---
title: "Advanced R - Chapter 1 - Data Structures"
author: "Matthew Strimas-Mackey"
output:
  html_document:
    keep_md: yes
---

# Data structures

## Quiz

1. What are the three properties of a vector, other than its contents?
- length
- type

1. What are the four common types of atomic vectors? What are the two rare types?
- character
- boolean
- numeric
- dimension

2. What are attributes? How do you get them and set them?

3. How is a list different from an atomic vector? How is a matrix different from a data frame?
- Only one datatype allowed in vector and matrix

4. Can you have a list that is a matrix? Can a data frame have a column that is a matrix?
- No

## Vectors

Vectores can be atomic vectors, i.e. all elements of the same type, or lists, with mixed types. They have 3 basic properties:

* Type, `typeof()`, what it is.
* Length, `length()`, how many elements it contains.
* Attributes, `attributes()`, additional arbitrary metadata.


```{r vectors}
v <- c(1,2,3)
names(v) <- c('one', 'two', 'three')
l <- list(v)

typeof(v)
length(v)
attributes(v)

# Testing identity
is.vector(v) # Vector with no attributes other than name
is.atomic(v) # Atomic vector
is.list(l) # List
is.atomic(v) || is.list(v) # Either type of vector
```

### Atomic vectors

Four common types: logical, integer, double (often called numeric), and character. Two rare types: complex and raw.

Atomic vectors are usually created with `c()`, short for combine:

```{r atomic}
# By default all numbers are stored as doubles
dbl_var <- c(1, 2.5, 4.5)
# With the L suffix, you get an integer rather than a double
int_var <- c(1L, 6L, 10L)

# Missing values specified with NA
miss <- c(1, 2, 3, NA)
# NA a logical by default, but always coerced to correct type, but can also specify explicitely
c(1, 2, 3, NA_real_)
c(1L, 2L, 3L, NA_integer_)
c('1', '2', '3', NA_character_)
```

#### Types and tests

Given a vector, you can determine its type with `typeof()`, or check if it's a specific type with an "is" function: `is.character()`, `is.double()`, `is.integer()`, `is.logical()`, or, more generally, `is.atomic()`. 

```{r type}
int_var <- c(1L, 6L, 10L)
typeof(int_var)
is.integer(int_var)
is.atomic(int_var)

dbl_var <- c(1, 2.5, 4.5)
typeof(dbl_var)
is.double(dbl_var)
is.atomic(dbl_var)

# is.numeric() returns T for double or integer
is.numeric(int_var)
is.numeric(dbl_var)
```

#### Coercion

When you attempt to combine different types they will be __coerced__ to the most flexible type. Types from least to most flexible are: logical, integer, double, and character. If confusion is likely, explicitly coerce with `as.character()`, `as.double()`, `as.integer()`, or `as.logical()`. 

### Lists

Lists are vectors whose elements can be of any type, including lists. ```list()``` creates a list out of it's elements, ```c()``` will combine lists together, ```unlist()``` turns a list into an atomic vector, making the required coercions.

```{r}
z <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(z)

# Difference between list and c
x <- list(list(1, 2), c(3, 4))
y <- c(list(1, 2), c(3, 4))
str(x)
str(y)

# unlist
str(unlist(x))
str(unlist(z)) # Coercion to string
```

### Exercises

1. What are the six types of atomic vector? How does a list differ from an atomic vector?
- double, integer, logical, character; complex, raw
- elements of a list can be of different types; atomic vectors are all of the same type


2. What makes `is.vector()` and `is.numeric()` fundamentally different to `is.list()` and `is.character()`?
- the first 2 will match 2 types of data, while the second will only match one each

3. Test your knowledge of vector coercion rules by predicting the output of
   the following uses of `c()`:

```{r}
c(1, FALSE) # c(1,0)
c("a", 1) # c('a', '1')
c(list(1), "a") # list(1, 'a')
c(TRUE, 1L) # c(1L, 1L)
```

4.  Why do you need to use `unlist()` to convert a list to an atomic vector? Why doesn't `as.vector()` work? 
- a list is already a vector

5. Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?
- Coercion

6. Why is the default missing value, `NA`, a logical vector? What's special about logical vectors? (Hint: think about `c(FALSE, NA_character_)`.)
- logical is the least flexible type

## Attributes

All objects can have arbitrary additional attributes, used to store metadata about the object. Attributes can be thought of as a named list (with unique names). Attributes can be accessed individually with `attr()` or all at once (as a list) with `attributes()`.

```{r}
y <- 1:10
names(y) <- 1:10
attr(y, "my_attribute") <- "This is a vector"
attr(y, "my_attribute")
str(attributes(y))
str(y)
```

#### Names

You can name a vector in three ways:

```{r}
# When creating it
c(a = 1, b = 2, c = 3)

# Modifying in place
x <- 1:3; names(x) <- c("a", "b", "c"); x
setNames(1:3, c("a", "b", "c"))
```

### Factors

A factor is a vector that can contain only predefined values, and is used to store categorical data. Factors are built on top of integer vectors using two attributes: the `class()`, "factor", which makes them behave differently from regular integer vectors, and the `levels()`, which defines the set of allowed values.

```{r}
x <- factor(c("a", "b", "b", "a"))
x
class(x)
levels(x)

# You can't use values that are not in the levels
x[2] <- "c"
x
```

Factors are useful when you know the possible values a variable may take, even if you don't see all values in a given dataset. Using a factor instead of a character vector makes it obvious when some groups contain no observations:

```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

### Exercises

1.  An early draft used this code to illustrate `structure()`:

    ```{r}
    structure(1:5, comment = "my attribute")
    ```

    But when you print that object you don't see the comment attribute.
    Why? Is the attribute missing, or is there something else special about
    it? (Hint: try using help.)
    
    - comment attributes are not printed by default
    
2.  What happens to a factor when you modify its levels? 
    
    ```{r}
    f1 <- factor(letters)
    levels(f1) <- rev(levels(f1))
    levels(f1) <- 1:26
    ```
    
    - the mapping between integers and levels is changed, so the vector is now labelled wrong

3.  What does this code do? How do `f2` and `f3` differ from `f1`?

    ```{r, results = "none"}
    f2 <- rev(factor(letters)); f2
    f3 <- factor(letters, levels = rev(letters)); f3
    ```
- First changes order of vector, second changes the order of the integers that are used to represent the levels

## Matrices and arrays

Adding a `dim()` attribute to an atomic vector allows it to behave like a multi-dimensional **array**. A special case of the array is the **matrix**, which has two dimensions. Matrices and arrays are created with `matrix()` and `array()`, or by using the assignment form of `dim()`:

```{r}
# Two scalar arguments to specify rows and columns
a <- matrix(1:6, ncol = 3, nrow = 2); a; class(a)
# One vector argument to describe all dimensions
b <- array(1:12, c(2, 3, 2)); b; class(b)

# You can also modify an object in place by setting dim()
c <- 1:6; dim(c) <- c(3, 2); c
dim(c) <- c(2, 3); c
```

`length()` and `names()` have high-dimensional generalisations:

* `length()` generalises to `nrow()` and `ncol()` for matrices, and `dim()`
  for arrays.

* `names()` generalises to `rownames()` and `colnames()` for matrices, and
  `dimnames()`, a list of character vectors, for arrays.

```{r}
length(a)
nrow(a)
ncol(a)
rownames(a) <- c("A", "B")
colnames(a) <- c("a", "b", "c")
a

length(b)
dim(b)
dimnames(b) <- list(c("one", "two"), c("a", "b", "c"), c("A", "B"))
b
```

### Exercises

1.  What does `dim()` return when applied to a vector?
    * NULL

2.  If `is.matrix(x)` is `TRUE`, what will `is.array(x)` return?
    * TRUE

3.  How would you describe the following three objects? What makes them
    different to `1:5`?

    ```{r}
    x1 <- array(1:5, c(1, 1, 5)); x1
    x2 <- array(1:5, c(1, 5, 1)); x2
    x3 <- array(1:5, c(5, 1, 1)); x3
    ```
    * 3d arrays, unlike 1:5 they have a dimension attribute

## Data frames

A data frame is a list of equal-length vectors. 

### Creation

You create a data frame using `data.frame()`, which takes named vectors as input:

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(df)

# Notice that, by default, strings were converted to factors. To suppress:
df <- data.frame(x = 1:3, y = c("a", "b", "c"), stringsAsFactors = F)
str(df)
```

### Testing and coercion

Because a `data.frame` is an S3 class, its type reflects the underlying vector used to build it: the list. To check if an object is a data frame, use `class()` or test explicitly with `is.data.frame()`:

```{r}
typeof(df)
class(df)
is.data.frame(df)
```

### Combining data frames

You can combine data frames using `cbind()` and `rbind()`:

```{r}
# Combine columns
cbind(df, data.frame(z = 3:1))
# Combine Rows
rbind(df, data.frame(x = 10, y = "z"))
```

It's a common mistake to try and create a data frame by `cbind()`ing vectors together. This doesn't work because `cbind()` will create a matrix unless one of the arguments is already a data frame. Instead use `data.frame()` directly:

```{r}
bad <- data.frame(cbind(a = 1:2, b = c("a", "b")))
str(bad)
good <- data.frame(a = 1:2, b = c("a", "b"),
  stringsAsFactors = FALSE)
str(good)
```

### Exercises

1.  What attributes does a data frame possess?
* names, rownames, class

1.  What does `as.matrix()` do when applied to a data frame with columns of different types?
* Coercion

1.  Can you have a data frame with 0 rows? What about 0 columns?
* yes