---
title: "Advanced R - Chapter 2 - Subsetting"
author: "Matthew Strimas-Mackey"
output:
  html_document:
    keep_md: yes
---


```{r, echo = FALSE}
rownames(mtcars) <- NULL
```

# Subsetting

##### Quiz

1.  What is the result of subsetting a vector with positive integers, 
    negative integers, a logical vector, or a character vector?

*   returns given elements, returns all but given elements, returns elements 
    corresponding to T, returns names elements 

2.  What's the difference between `[`, `[[`, and `$` when applied to a list?

*   `[` returns a list, `[[` a vector, and $ also a vector by names of elements

3.  When should you use `drop = FALSE`?

*   ??

4.  If `x` is a matrix, what does `x[] <- 0` do? How is it different to
    `x <- 0`?
    
*   Sets all elements to 0, sets it to a vector of length 1 with value 0 

5.  How can you use a named vector to relabel categorical variables?

*   ??

## Data types {#data-types}

Six data types can be used for subsetting.

### Atomic vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)

# 1. Positive integers: return elements
x[c(1,3,1)]

# 2. Negative integers: omit elements
x[-c(2, 4)]

# 3. Logical vectors: select elements that meet a condition
x[c(T, F, T, F)]
x[x>3]

# 4. Nothing: returns whole vector
x[]

# 5. Zero: returns zero-length vector
x[0]

# 6. Characters: returns specified named elements
y <- setNames(x, letters[1:4])
y[c('a', 'd')]
```

### Lists

Subsetting a list works in the same way as subsetting an atomic vector. Using `[` will always return a list; `[[` and `$`, as described below, let you pull out the components of the list. 

### Matrices and arrays {#matrix-subsetting}

You can subset higher-dimensional structures in three ways:

* With multiple vectors.

```{r}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")
rownames(a) <- c("D", "E", "F")
a[1:2, ]
a[c(T, F, T), c("B", "A")]
a[0, -2]
```

* With a single vector: arrays are vectors with dimension, columns filled first

```{r}
(vals <- outer(1:5, 1:5, FUN = "paste", sep = ","))
vals[c(4, 15)]
```

*   With a matrix: rows correspond to elements you want to select and columns
    correspond to the coordinates

```{r}
vals <- outer(1:5, 1:5, FUN = "paste", sep = ",")
select <- matrix(ncol = 2, byrow = TRUE, c(
  1, 1,
  3, 1,
  2, 4
))
vals[select]
```

### Data frames

Data frames possess the characteristics of both lists and matrices: if you subset with a single vector, they behave like lists; if you subset with two vectors, they behave like matrices.

```{r}
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])

# Rows are subset like matrices
df[df$x == 2, ]
df[c(1, 3), ]

# There are two ways to select columns from a data frame
# Like a list:
df[c("x", "z")]
# Like a matrix
df[, c("x", "z")]

# There's an important difference if you select a single 
# column: matrix subsetting simplifies by default, list 
# subsetting does not.
str(df["x"])
str(df[, "x"])
```

### Exercises

1.  Fix each of the following common data frame subsetting errors:

    ```{r, eval = FALSE}
    mtcars[mtcars$cyl == 4, ]
    mtcars[-(1:4), ]
    mtcars[mtcars$cyl <= 5, ]
    mtcars[mtcars$cyl %in% c(4,6), ]
    ```

1.  Why does `x <- 1:5; x[NA]` yield five missing values? (Hint: why is 
    it different from `x[NA_real_]`?)

*   R thinks NA is logical and recycles it

1.  What does `upper.tri()` return? How does subsetting a matrix with it 
    work? Do we need any additional subsetting rules to describe its behaviour?

    ```{r, eval = FALSE}
    x <- outer(1:5, 1:5, FUN = "*")
    x[upper.tri(x)]
    ```
    
*   A matrix of same dimension as x that has T in upper triangle and F elsewhere

1.  Why does `mtcars[1:20]` return a error? How does it differ from the 
    similar `mtcars[1:20, ]`?
    
*   Tries to return columns 1:20, but there are only 11

1.  Implement your own function that extracts the diagonal entries from a
    matrix (it should behave like `diag(x)` where `x` is a matrix).
    
    ```{r myDiag}
    myDiag <- function(x) {
        i <- min(dim(x))
        x[matrix(c(1:i,1:i), ncol=2)]
    }
    ```

1.  What does `df[is.na(df)] <- 0` do? How does it work?

* Changes 0's to NAs

## Subsetting operators {#subsetting-operators}

>  "If list `x` is a train carrying objects, then `x[[5]]` is
> the object in car 5; `x[4:6]` is a train of cars 4-6." 
>
> --- @RLangTip

```{r}
a <- list(a = 1, b = 2)
a[1]; a[[1]]
a["a"]; a[["a"]]

# If you do supply a vector it indexes recursively
b <- list(a = list(b = list(c = list(d = 1))))
b[[c("a", "b", "c", "d")]]
# Same as
b[["a"]][["b"]][["c"]][["d"]]
```

### Simplifying vs. preserving subsetting

Simplifying subsets returns the simplest possible data structure that can represent the output, and is useful interactively because it usually gives you what you want. Preserving subsetting keeps the structure of the output the same as the input, and is generally better for programming because the result will always be the same type.

|             | Simplifying               | Preserving                                   |
|-------------|---------------------------|----------------------------------------------|
| Vector      | `x[[1]]`                  | `x[1]`                                       |
| List        | `x[[1]]`                  | `x[1]`                                       |
| Factor      | `x[1:4, drop = T]`        | `x[1:4]`                                     |
| Array       | `x[1, ]` __or__ `x[, 1]`  | `x[1, , drop = F]` __or__ `x[, 1, drop = F]` |
| Data frame  | `x[, 1]` __or__ `x[[1]]`  | `x[, 1, drop = F]` __or__ `x[1]`             |

Preserving is the same for all data types: you get the same type of output as input. Simplifying behaviour varies slightly between different data types, as described below:

*   __Atomic vector__: removes names.

    ```{r}
    x <- c(a = 1, b = 2)
    x[1]
    x[[1]]
    ```

*   __List__: return the object inside the list, not a single element list.

    ```{r}
    y <- list(a = 1, b = 2)
    str(y[1])
    str(y[[1]])
    ```

*   __Factor__: drops any unused levels.

    ```{r}
    z <- factor(c("a", "b"))
    z[1]
    z[1, drop = TRUE]
    ```

*   __Matrix__ or __array__: if any of the dimensions has length 1, 
    drops that dimension.

    ```{r}
    a <- matrix(1:4, nrow = 2)
    a[1, , drop = FALSE]
    a[1, ]
    ```

*   __Data frame__: if output is a single column, returns a vector instead of 
    a data frame.

    ```{r}
    (df <- data.frame(a = 1:2, b = 1:2))
    str(df[1])
    str(df[[1]])
    str(df[, "a", drop = FALSE])
    str(df[, "a"])
    ```

### `$`

`$` is a shorthand operator, where `x$y` is equivalent to `x[["y", exact = FALSE]]`, i.e. it does partial matching.

### Missing/out of bounds indices

The following table summarises the results of subsetting atomic vectors and lists with `[` and `[[` and different types of OOB value.

| Operator | Index       | Atomic      | List          |
|----------|-------------|-------------|---------------|
| `[`      | OOB         | `NA`        | `list(NULL)`  |
| `[`      | `NA_real_`  | `NA`        | `list(NULL)`  |
| `[`      | `NULL`      | `x[0]`      | `list(NULL)`  |
| `[[`     | OOB         | Error       | Error         |
| `[[`     | `NA_real_`  | Error       | `NULL`        |
| `[[`     | `NULL`      | Error       | Error         |

### Exercises

1.  Given a linear model, e.g., `mod <- lm(mpg ~ wt, data = mtcars)`, extract
    the residual degrees of freedom. Extract the R squared from the model
    summary (`summary(mod)`)

    ```{r}
    mod <- lm(mpg ~ wt, data = mtcars)
    mod$df.residual
    summary(mod)$r.squared
    ```

## Subsetting and assignment

All subsetting operators can be combined with assignment to modify selected values of the input vector.

Subsetting with nothing can be useful in conjunction with assignment because it will preserve the original object class and structure. Compare the following two expressions. In the first, `mtcars` will remain as a data frame. In the second, `mtcars` will become a list.

```{r, eval = FALSE}
mtcars[] <- sapply(mtcars, as.integer)
mtcars <- lapply(mtcars, as.integer)
```

With lists, you can use subsetting + assignment + `NULL` to remove components from a list. To add a literal `NULL` to a list, use `[` and `list(NULL)`:

```{r}
x <- list(a = 1, b = 2)
x[["b"]] <- NULL
str(x)

y <- list(a = 1)
y["b"] <- list(NULL)
str(y)
```

## Applications

### Lookup tables (character subsetting)

Character matching provides a powerful way to make lookup tables. Say you want to convert abbreviations:

```{r}
x <- c("m", "f", "u", "f", "f", "m", "m")
lookup <- c(m = "Male", f = "Female", u = NA)
lookup[x]
unname(lookup[x])

# Or with fewer output values
c(m = "Known", f = "Known", u = "Unknown")[x]
```

### Matching and merging by hand (integer subsetting) {#matching-merging}

You may have a more complicated lookup table which has multiple columns of information. Suppose we have a vector of integer grades, and a table that describes their properties:

```{r}
grades <- c(1, 2, 2, 3, 1)

info <- data.frame(
  grade = 3:1,
  desc = c("Excellent", "Good", "Poor"),
  fail = c(F, F, T)
)
```

We want to duplicate the info table so that we have a row for each value in `grades`. We can do this in two ways, either using `match()` and integer subsetting, or `rownames()` and character subsetting:

```{r}
grades

# Using match
id <- match(grades, info$grade)
info[id, ]

# Using rownames
rownames(info) <- info$grade
info[as.character(grades), ]
```

### Random samples/bootstrap (integer subsetting)

You can use integer indices to perform random sampling or bootstrapping of a vector or data frame. `sample()` generates a vector of indices, then subsetting to access the values:

```{r}
df <- data.frame(x = rep(1:3, each = 2), y = 6:1, z = letters[1:6])

# Randomly reorder
df[sample(nrow(df)), ]
# Select 3 random rows
df[sample(nrow(df), 3), ]
# Select 6 bootstrap replicates
df[sample(nrow(df), 6, rep = T), ]
```

### Ordering (integer subsetting)

`order()` takes a vector as input and returns an integer vector describing how the subsetted vector should be ordered:

```{r}
x <- c("b", "c", "a")
order(x)
x[order(x)]
```

For two or more dimensions, `order()` and integer subsetting makes it easy to order either the rows or columns of an object:

```{r}
# Randomly reorder df
df2 <- df[sample(nrow(df)), 3:1]
df2

df2[order(df2$x), ]
df2[, order(names(df2))]
```

### Expanding aggregated counts (integer subsetting)

Sometimes you get a data frame where identical rows have been collapsed into one and a count column has been added. `rep()` and integer subsetting make it easy to uncollapse the data by subsetting with a repeated row index:

```{r}
df <- data.frame(x = c(2, 4, 1), y = c(9, 11, 6), n = c(3, 5, 1))
rep(1:nrow(df), df$n)
df3 <- df[rep(1:nrow(df), df$n), ]
(rownames(df3) <- NULL)
```

### Removing columns from data frames (character subsetting)

There are two ways to remove columns from a data frame. You can set individual columns to NULL: 

```{r}
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
df$z <- NULL
```

Or you can subset to return only the columns you want:

```{r}
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
df[c("x", "y")]
```

If you know the columns you don't want, use set operations to work out which colums to keep:

```{r}
df[setdiff(names(df), "z")]
```

### Selecting rows based on a condition (logical subsetting)

Because it allows you to easily combine conditions from multiple columns, logical subsetting is probably the most commonly used technique for extracting rows out of a data frame. 

```{r}
mtcars[mtcars$gear == 5, ]
mtcars[mtcars$gear == 5 & mtcars$cyl == 4, ]
```

Remember to use the vector boolean operators `&` and `|`, not the short-circuiting scalar operators `&&` and `||` which are more useful inside if statements. Don't forget [De Morgan's laws][demorgans], which can be useful to simplify negations:

* `!(X & Y)` is the same as `!X | !Y`
* `!(X | Y)` is the same as `!X & !Y`

For example, `!(X & !(Y | Z))` simplifies to `!X | !!(Y|Z)`, and then to `!X | Y | Z`.

### Boolean algebra vs. sets (logical & integer subsetting)

It's useful to be aware of the natural equivalence between set operations (integer subsetting) and boolean algebra (logical subsetting). Using set operations is more effective when:

* You want to find the first (or last) `TRUE`.

* You have very few `TRUE`s and very many `FALSE`s; a set representation 
  may be faster and require less storage.

`which()` allows you to convert a boolean representation to an integer representation. There's no reverse operation in base R but we can easily create one: \indexc{which()}

```{r}
x <- sample(10) < 4
which(x)

unwhich <- function(x, n) {
  out <- rep_len(FALSE, n)
  out[x] <- TRUE
  out
}
unwhich(which(x), 10)
```

Let's create two logical vectors and their integer equivalents and then explore the relationship between boolean and set operations.

```{r}
(x1 <- 1:10 %% 2 == 0)
(x2 <- which(x1))
(y1 <- 1:10 %% 5 == 0)
(y2 <- which(y1))

# X & Y <-> intersect(x, y)
x1 & y1
intersect(x2, y2)

# X | Y <-> union(x, y)
x1 | y1
union(x2, y2)

# X & !Y <-> setdiff(x, y)
x1 & !y1
setdiff(x2, y2)

# xor(X, Y) <-> setdiff(union(x, y), intersect(x, y))
xor(x1, y1)
setdiff(union(x2, y2), intersect(x2, y2))
```

### Exercises

1.  How would you randomly permute the columns of a data frame? (This is an
    important technique in random forests.) Can you simultaneously permute 
    the rows and columns in one step?
    
    ```{r}
    (df <- data.frame(x = 1:6, y = 6:1, z = letters[1:6]))
    df[sample(nrow(df)), sample(ncol(df))]
    ```

1.  How would you select a random sample of `m` rows from a data frame? 
    What if the sample had to be contiguous (i.e., with an initial row, a 
    final row, and every row in between)?
    
    ```{r}
    (df <- data.frame(x = 1:6, y = 6:1, z = letters[1:6]))
    m <- 4
    df[seq(sample(nrow(df) - m + 1, 1), length.out=m), ]
    ```
    
1.  How could you put the columns in a data frame in alphabetical order?

    ```{r}
    (df <- data.frame(x = 1:6, a = 6:1, l = letters[1:6]))
    df[, order(names(df))]
    ```

## Answers

1.  Positive integers select elements at specific positions, negative integers
    drop elements; logical vectors keep elements at positions corresponding to
    `TRUE`; character vectors select elements with matching names.
   
1.  `[` selects sub-lists. It always returns a list; if you use it with a
    single positive integer, it returns a list of length one. `[[` selects 
    an element within a list. `$` is a convenient shorthand: `x$y` is
    equivalent to `x[["y"]]`.

1.  Use `drop = FALSE` if you are subsetting a matrix, array, or data frame 
    and you want to preserve the original dimensions. You should almost 
    always use it when subsetting inside a function.
   
1.  If `x` is a matrix, `x[] <- 0` will replace every element with 0, 
    keeping the same number of rows and columns. `x <- 0` completely 
    replaces the matrix with the value 0.
    
1.  A named character vector can act as a simple lookup table: 
    `c(x = 1, y = 2, z = 3)[c("y", "z", "x")]`