---
title: "Advanced R - Chapter 4 - Functions"
author: "Matthew Strimas-Mackey"
output:
  html_document:
    keep_md: yes
---

# Functions

Functions are a fundamental building block of R: to master many of the more advanced techniques in this book, you need a solid foundation in how functions work. You've probably already created many R functions, and you're familiar with the basics of how they work. The focus of this chapter is to turn your existing, informal knowledge of functions into a rigorous understanding of what functions are and how they work. You'll see some interesting tricks and techniques in this chapter, but most of what you'll learn will be more important as the building blocks for more advanced techniques.

The most important thing to understand about R is that functions are objects in their own right. You can work with them exactly the same way you work with any other type of object. This theme will be explored in depth in functional programming.

##### Quiz

Answer the following questions to see if you can safely skip this chapter. You can find the answers at the end of the chapter in [answers](#function-answers).

1.  What are the three components of a function?

    *   parameters, return statement, ?

1.  What does the following code return?

    ```{r, eval = FALSE}
    x <- 10
    f1 <- function(x) {
      function() {
        x + 10
      }
    }
    f1(1)()
    ```
    
    *   11
    
1.  How would you more typically write this code?

    ```{r, eval = FALSE}
    `+`(1, `*`(2, 3))
    ```
    
    *   `1 + (2 * 3)`
    
1.  How could you make this call easier to read?

    ```{r, eval = FALSE}
    mean(, TRUE, x = c(1:10, NA))
    ```
    
    *   `mean(c(1:10, NA), na.rm=T)`

1.  Does the following function throw an error when called? Why/why not?

    ```{r, eval = FALSE}
    f2 <- function(a, b) {
      a * 10
    }
    f2(10, stop("This is an error!"))
    ```

    *   No
    
1.  What is an infix function? How do you write it? What's a replacement 
    function? How do you write it?
    
    *   ?

1.  What function do you use to ensure that a cleanup action occurs 
    regardless of how a function terminates?
    
    *   `on.exit()`

## Function components

All R functions have three parts: 

* the `body()`, the code inside the function.

* the `formals()`, the list of arguments which controls how you can call the function.

* the `environment()`, the "map" of the location of the function's variables.

When you print a function in R, it shows you these three important components. If the environment isn't displayed, it means that the function was created in the global environment. 

```{r}
f <- function(x) x^2
f

formals(f)
body(f)
environment(f)
```

The assignment forms of `body()`, `formals()`, and `environment()` can also be used to modify functions.

Like all objects in R, functions can also possess any number of additional `attributes()`. 

### Primitive functions

There is one exception to the rule that functions have three components. Primitive functions, like `sum()`, call C code directly with `.Primitive()` and contain no R code. Therefore their `formals()`, `body()`, and `environment()` are all `NULL`.

### Exercises

1.  What function allows you to tell if an object is a function? What function
    allows you to tell if a function is a primitive function?
    
    *   `is.function()` and `is.primitive()`

1.  This code makes a list of all functions in the base package. 
    
    ```{r}
    objs <- mget(ls("package:base"), inherits = TRUE)
    funs <- Filter(is.function, objs)
    ```

    Use it to answer the following questions:

    a. Which base function has the most arguments?
    
    ```{r}
    which.max(sapply(funs, FUN=function(x) length(formals(x))))
    ```
    
    a. How many base functions have no arguments? What's special about those
       functions?
    
    ```{r}
    # These are all primitives
    sum(sapply(funs, FUN=function(x) length(formals(x)))==0)
    ```
       
    a. How could you adapt the code to find all primitive functions?
    
    ```{r}
    Filter(is.primitive, objs)
    ```

1. What are the three important components of a function?

    * formals, body, environment

1. When does printing a function not show what environment it was created in?
    
    * For primitive functions

## Lexical scoping

Scoping is the set of rules that govern how R looks up the value of a symbol.

R has two types of scoping: __lexical scoping__, implemented automatically at the language level, and __dynamic scoping__, used in select functions to save typing during interactive analysis.

Lexical scoping looks up symbol values based on how functions were nested when they were created, not how they are nested when they are called. With lexical scoping, you don't need to know how the function is called to figure out where the value of a variable will be looked up. You just need to look at the function's definition.

There are four basic principles behind R's implementation of lexical scoping:

### Name masking

All functions have an evaluation environment, created when the function is called, and an enclosing environment, the environment active when the fuction was created. When variables are references, R first searches the evaluation environment, then the enclosing environment recursively, all the way up to the global environment, then down through all the loaded packages.

```{r, eval = FALSE}
x <- 1
h <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z)
  }
  i()
}
h()
rm(x, h)
```

The same rules apply to closures, functions created by other functions. Closures include the environment in which the function was created. This allows access to the variables in this envlosing environment

```{r, eval = FALSE}
j <- function(x) {
  y <- 2
  function() {
    c(x, y)
  }
}
k <- j(1)
k()
environment(k); environment(j)
rm(j, k)
```

### Functions vs. variables

The same principles apply regardless of the type of associated value --- finding functions works exactly the same way as finding variables, except if you are using a name in a context where it's obvious that you want a function (e.g., `f(3)`), R will ignore objects that are not functions while it is searching.

```{r}
n <- function(x) x / 2
o <- function() {
  n <- 10
  n(n)
}
o()
rm(n, o)
```

### A fresh start

Every time a function is called, a new environment is created to host execution.

```{r, eval = FALSE}
j <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  print(a)
}
j()
rm(j)
```

### Dynamic lookup

Lexical scoping determines where to look for values, not when to look for them. R looks for values when the function is run, not when it's created. This means that the output of a function can be different depending on objects outside its environment: 

```{r}
f <- function() x
x <- 15
f()

x <- 20
f()
```

You generally want to avoid this behaviour because it means the function is no longer self-contained. This is a common error --- if you make a spelling mistake in your code, you won't get an error when you create the function, and you might not even get one when you run the function, depending on what variables are defined in the global environment.

One way to detect this problem is the `findGlobals()` function from `codetools`. This function lists all the external dependencies of a function:

```{r}
f <- function() x + 1
codetools::findGlobals(f)
```

### Exercises

1. What does the following code return? Why? What does each of the three `c`'s mean?

    ```{r, eval = FALSE}
    c <- 10
    c(c = c)
    ```
    
    *   c refers to a variable, a formal, and a function. Returns 10.

2. What are the four principles that govern how R looks for values?

    * Lexical scoping with dynamic lookup

3. What does the following function return? Make a prediction before 
   running the code yourself.

    ```{r, eval = FALSE}
    f <- function(x) {
      f <- function(x) {
        f <- function(x) {
          x ^ 2
        }
        f(x) + 1
      }
      f(x) * 2
    }
    f(10)
    ```

## Every operation is a function call

> "To understand computations in R, two slogans are helpful:
>
> * Everything that exists is an object.
> * Everything that happens is a function call."
>
> --- John Chambers

Every operation in R is a function call, including the use of infix operators. Note that `` ` ``, the backtick, lets you refer to functions or variables that have otherwise reserved or illegal names: 

```{r}
x <- 10; y <- 5
x + y
`+`(x, y)

for (i in 1:2) print(i)
`for`(i, 1:2, print(i))

if (i == 1) print("yes!") else print("no.")
`if`(i == 1, print("yes!"), print("no."))

x[3]
`[`(x, 3)

{ print(1); print(2); print(3) }
`{`(print(1), print(2), print(3))
```

It is possible to override the definitions of these special functions, For example, this feature makes it possible for the `dplyr` package to translate R expressions into SQL expressions. Domain specific languages uses this idea to create domain specific languages that allow you to concisely express new concepts using existing R constructs.

It's more often useful to treat special functions as ordinary functions. For example, we could use `sapply()` to add 3 to every element of a list by first using the built-in `+` function:

```{r}
sapply(1:5, `+`, 3)
sapply(1:5, "+", 3)
```

Note the difference between `` `+` `` and `"+"`.  The first one is the value of the object called `+`, and the second is a string containing the character `+`.  The second version works because `lapply` can be given the name of a function instead of the function itself: if you read the source of `lapply()`, you'll see the first line uses `match.fun()` to find functions given their names.

A more useful application is to combine `lapply()` or `sapply()` with subsetting:

```{r}
x <- list(1:3, 4:9, 10:12)
sapply(x, `[`, 2)

# equivalent to
sapply(x, function(x) x[2])
```

## Function arguments 

It's useful to distinguish between the formal arguments and the actual arguments of a function. The formal arguments are a property of the function, whereas the actual or calling arguments can vary each time you call the function.

### Calling functions

When calling a function you can specify arguments by position, by complete name, or by partial name. Arguments are matched first by exact name (perfect matching), then by prefix matching, and finally by position. 

```{r, error = TRUE}
f <- function(abcdef, bcde1, bcde2) {
  list(a = abcdef, b1 = bcde1, b2 = bcde2)
}
str(f(1, 2, 3))
str(f(2, 3, abcdef = 1))

# Can abbreviate long argument names:
str(f(2, 3, a = 1))

# But this doesn't work because abbreviation is ambiguous
str(f(1, 3, b = 1))
```

### Calling a function given a list of arguments

Suppose you had a list of function arguments: 

```{r}
args <- list(1:10, na.rm = TRUE)
```

How could you then send that list to `mean()`?  You need `do.call()`:

```{r}
do.call(mean, list(1:10, na.rm = TRUE))
# Equivalent to
mean(1:10, na.rm = TRUE)
```

### Default and missing arguments

Function arguments in R can have default values. 

```{r}
f <- function(a = 1, b = 2) {
  c(a, b)
}
f()
```

Default value can be defined in terms of other arguments:

```{r}
g <- function(a = 1, b = a * 2) {
  c(a, b)
}
g()
g(10)
```

Default arguments can even be defined in terms of variables created within the function. 

```{r}
h <- function(a = 1, b = d) {
  d <- (a + 1) ^ 2
  c(a, b)
}
h()
h(10)
```

You can determine if an argument was supplied or not with the `missing()` function. 

```{r}
i <- function(a, b) {
  c(missing(a), missing(b))
}
i()
i(a = 1)
i(b = 2)
i(1, 2)
```

Sometimes you want to add a non-trivial default value, which might take several lines of code to compute. Instead of inserting that code in the function definition, you could use `missing()` to conditionally compute it if needed. However, this makes it hard to know which arguments are required and which are optional without carefully reading the documentation. Instead, I usually set the default value to `NULL` and use `is.null()` to check if the argument was supplied.

### Lazy evaluation

By default, R function arguments are lazy --- they're only evaluated if they're actually used: 

```{r}
f <- function(x) {
  10
}
f(stop("This is an error!"))
```

If you want to ensure that an argument is evaluated you can use `force()`: 

```{r, error = TRUE}
f <- function(x) {
  force(x)
  10
}
f(stop("This is an error!"))
```

This is important when creating closures with `lapply()` or a loop:

```{r}
add <- function(x) {
  function(y) x + y
}
adders <- lapply(1:10, add)
adders[[1]](10)
adders[[10]](10)
```

`x` is lazily evaluated the first time that you call one of the adder functions. At this point, the loop is complete and the final value of `x` is 10.  Therefore all of the adder functions will add 10 on to their input, probably not what you wanted!  Manually forcing evaluation fixes the problem:

```{r}
add <- function(x) {
  force(x)
  function(y) x + y
}
adders2 <- lapply(1:10, add)
adders2[[1]](10)
adders2[[10]](10)
```

This behaviour occurs because, when a function is called, its formal arguments are bound to promise objects, which store the expression given for the formal argument and a pointer to the environment that the function was called from. Only when the argument is accessed is the expression evaluated in the stored environment and the value stored within the promise.

```{r}
require(pryr)
require(dplyr)
    
addpromise <- function(x) {
  print(promise_info(x))
  function(y) x + y
}

# Note that all promises for x have the same environment, i.e. the evaluation environment of the lapply call
adders <- lapply(1:10, addpromise)
adders[[1]](10)
```


Default arguments are evaluated inside the function. This means that if the expression depends on the current environment the results will differ depending on whether you use the default value or explicitly provide one.

```{r}
f <- function(x = ls()) {
  a <- 1
  x
}

# ls() evaluated inside f:
f()

# ls() evaluated in global environment:
f(ls())
```

An unevaluated argument is called a __promise__. A promise is made up of two parts: 

* The expression which gives rise to the delayed computation. 

* The environment where the expression was created and where it should be 
  evaluated.

The first time a promise is accessed the expression is evaluated in the environment where it was created. This value is cached, so that subsequent access to the evaluated promise does not recompute the value (but the original expression is still associated with the value). You can find more information about a promise using `pryr::promise_info()`. This uses some C++ code to extract information about the promise without evaluating it, which is impossible to do in pure R code.

### `...`

There is a special argument called `...` .  This argument will match any arguments not otherwise matched, and can be easily passed on to other functions.  This is useful if you want to collect arguments to call another function, but you don't want to prespecify their possible names. 

To capture `...` in a form that is easier to work with, you can use `list(...)`.

```{r}
f <- function(...) {
  names(list(...))
}
f(a = 1, b = 2)
```

Using `...` comes at a price --- any misspelled arguments will not raise an error, and any arguments after `...` must be fully named.  This makes it easy for typos to go unnoticed:

```{r}
sum(1, 2, NA, na.mr = TRUE)
```

### Exercises

1.  Clarify the following list of odd function calls:

    ```{r, eval = FALSE}
    x <- sample(replace = TRUE, 20, x = c(1:10, NA)) # size=20, matched by order
    y <- runif(min = 0, max = 1, 20) # n=20, matched by order after others matched by name
    cor(m = "k", y = y, u = "p", x = x) # partial matching m=method, u=use
    ```

1.  What does this function return? Why? Which principle does it illustrate?
  
    ```{r, eval = FALSE}    
    f1 <- function(x = {y <- 1; 2}, y = 0) {
      x + y
    }
    f1()
    ```
    
    *   3; lazy evaluation, dynamic lookup

1.  What does this function return? Why? Which principle does it illustrate?

    ```{r, eval = FALSE}
    f2 <- function(x = z) {
      z <- 100
      x
    }
    f2()
    ```
    
    *   100; dynamic lookup

## Special calls 

R supports two additional syntaxes for calling special types of functions: infix and replacement functions.

### Infix functions {#infix-functions}

Most functions in R are "prefix" operators: the name of the function comes before the arguments. You can also create infix functions where the function name comes in between its arguments, like `+` or `-`.  All user created infix functions must start and end with `%` and R comes with the following infix functions predefined: `%%`, `%*%`, `%/%`, `%in%`, `%o%`,  `%x%`. (The complete list of built-in infix operators that don't need `%` is: `::, :::, $, @, ^, *, /, +, -, >, >=, <, <=, ==, !=, !, &, &&, |, ||, ~, <-, <<-`) 

For example, we could create a new operator that pastes together strings:

```{r}
`%+%` <- function(a, b) paste(a, b, sep = "")
"new" %+% " string"
```

Note that when creating the function, you have to put the name in backticks because it's a special name.

The names of infix functions are more flexible than regular R functions: they can contain any sequence of characters (except "%", of course). You will need to escape any special characters in the string used to define the function, but not when you call it:

```{r}
`% %` <- function(a, b) paste(a, b)
`%'%` <- function(a, b) paste(a, b)
`%/\\%` <- function(a, b) paste(a, b)

"a" % % "b"
"a" %'% "b"
"a" %/\% "b"
```

R's default precedence rules mean that infix operators are composed from left to right:

```{r}
`%-%` <- function(a, b) paste0("(", a, " %-% ", b, ")")
"a" %-% "b" %-% "c"
```

### Replacement functions 

Replacement functions act like they modify their arguments in place, and have the special name `xxx<-`. They typically have two arguments (`x` and `value`), although they can have more, and they must return the modified object. For example, the following function allows you to modify the second element of a vector: 

```{r}
`second<-` <- function(x, value) {
  x[2] <- value
  x
}
x <- 1:10
second(x) <- 5L
x
```

When R evaluates the assignment `second(x) <- 5`, it notices that the left hand side of the `<-` is not a simple name, so it looks for a function named `second<-` to do the replacement. 

I say they "act" like they modify their arguments in place, because they actually create a modified copy. We can see that by using `pryr::address()` to find the memory address of the underlying object.

```{r, message = FALSE}
library(pryr)
x <- 1:10
address(x)
second(x) <- 6L
address(x)
```

If you want to supply additional arguments, they go in between `x` and `value`:

```{r}
`modify<-` <- function(x, position, value) {
  x[position] <- value
  x
}
modify(x, 1) <- 10
x
```

When you call `modify(x, 1) <- 10`, behind the scenes R turns it into:

```{r, eval = FALSE}
x <- `modify<-`(x, 1, 10)
```

It's often useful to combine replacement and subsetting:

```{r}
x <- c(a = 1, b = 2, c = 3)
names(x)
names(x)[2] <- "two"
names(x)
```

This works because the expression `names(x)[2] <- "two"` is evaluated as if you had written:

```{r, eval = FALSE}
`*tmp*` <- names(x)
`*tmp*`[2] <- "two"
names(x) <- `*tmp*`
```

(Yes, it really does create a local variable named `*tmp*`, which is removed afterwards.)

### Exercises

1. Create a list of all the replacement functions found in the base package. 
   Which ones are primitive functions?
   
    ```{r}
    pkg <- ls("package:base")[grep('<-', ls("package:base"))]
    objs <- mget(pkg, inherits = TRUE)
    funs <- Filter(is.primitive, objs)
    ```

2. What are valid names for user created infix functions?

    *   must be of the form `%X%`, where X is user defined

3. Create an infix `xor()` operator.

    ```{r}
    `%xor%` <- function(x, y) xor(x,y)
    T %xor% F
    F %xor% T
    F %xor% F
    T %xor% T
    ```

4. Create infix versions of the set functions `intersect()`, `union()`, and 
   `setdiff()`.

5. Create a replacement function that modifies a random location in a vector.

    ```{r}
    `repRand<-` <- function(x,value) {
        x[sample(1:length(x), 1)] <- value
        x
    }
    x <- 1:65
    repRand(x) <- 987
    x
    ```

## Return values 

The last expression evaluated in a function becomes the return value, the result of invoking the function. Generally, I think it's good style to reserve the use of an explicit `return()` for when you are returning early, such as for an error, or a simple case of the function. 

Functions can return only a single object. But this is not a limitation because you can return a list containing any number of objects.

The functions that are the easiest to understand and reason about are pure functions: functions that always map the same input to the same output and have no other impact on the workspace. In other words, pure functions have no __side effects__: they don't affect the state of the world in any way apart from the value they return. 

R protects you from one type of side effect: most R objects have copy-on-modify semantics. So modifying a function argument does not change the original value.

Most base R functions are pure, with a few notable exceptions:

* `library()` which loads a package, and hence modifies the search path.

* `setwd()`, `Sys.setenv()`, `Sys.setlocale()` which change the working 
  directory, environment variables, and the locale, respectively.

* `plot()` and friends which produce graphical output.

* `write()`, `write.csv()`, `saveRDS()`, etc. which save output to disk.

* `options()` and `par()` which modify global settings.

* S4 related functions which modify global tables of classes and methods.

* Random number generators which produce different numbers each time you 
  run them.

It's generally a good idea to minimise the use of side-effects, and where possible, to minimise the footprint of side effects by separating pure from impure functions. Pure functions are easier to test (because all you need to worry about are the input values and the output), and are less likely to work differently on different versions of R or on different platforms. 

Functions can return `invisible` values, which are not printed out by default when you call the function. 

```{r}
f1 <- function() 1
f2 <- function() invisible(1)

f1()
f2()
f1() == 1
f2() == 1
```

You can force an invisible value to be displayed by wrapping it in parentheses:

```{r}
(f2())
```

The most common function that returns invisibly is `<-`: 

```{r}
a <- 2
(a <- 2)
```

### On exit 

As well as returning a value, functions can set up other triggers to occur when the function is finished using `on.exit()`. This is often used as a way to guarantee that changes to the global state are restored when the function exits. The code in `on.exit()` is run regardless of how the function exits, whether with an explicit (early) return, an error, or simply reaching the end of the function body. 

```{r}
in_dir <- function(dir, code) {
  old <- setwd(dir)
  on.exit(setwd(old))

  force(code)
}
getwd()
in_dir("~", getwd())
```

The basic pattern is simple:

* We first set the directory to a new location, capturing the current location 
  from the output of `setwd()`.

* We then use `on.exit()` to ensure that the working directory is returned to 
  the previous value regardless of how the function exits.

* Finally, we explicitly force evaluation of the code. (We don't actually need 
  `force()` here, but it makes it clear to readers what we're doing.)

**Caution**: If you're using multiple `on.exit()` calls within a function, make sure to set `add = TRUE`. Unfortunately, the default in `on.exit()` is `add = FALSE`, so that every time you run it, it overwrites existing exit expressions. Because of the way `on.exit()` is implemented, it's not possible to create a variant with `add = TRUE`, so you must be careful when using it.

### Exercises

1.  How does the `chdir` parameter of `source()` compare to `in_dir()`? Why 
    might you prefer one approach to the other?

1.  What function undoes the action of `library()`? How do you save and restore
    the values of `options()` and `par()`?
    
    ```{r}
    detach()
    oldOptions <- options("Changes go here"); options(oldOptions)
    oldPar <- par("Changes go here"); par(oldPar)
    ```

1.  Write a function that opens a graphics device, runs the supplied code, and 
    closes the graphics device (always, regardless of whether or not the plotting 
    code worked).

1.  We can use `on.exit()` to implement a simple version of `capture.output()`.

    ```{r}
    capture.output2 <- function(code) {
      temp <- tempfile()
      on.exit(file.remove(temp), add = TRUE)

      sink(temp)
      on.exit(sink(), add = TRUE)

      force(code)
      readLines(temp)
    }
    capture.output2(cat("a", "b", "c", sep = "\n"))
    ```

    Compare `capture.output()` to `capture.output2()`. How do the functions 
    differ? What features have I removed to make the key ideas easier to see? 
    How have I rewritten the key ideas to be easier to understand?


## Quiz answers {#function-answers}

\enlargethispage*{\baselineskip}

1.  The three components of a function are its body, arguments, and environment.

1.  `f1(1)()` returns 11.

1.  You'd normally write it in infix style: `1 + (2 * 3)`.

1.  Rewriting the call to `mean(c(1:10, NA), na.rm = TRUE)` is easier to
    understand.
    
1.  No, it does not throw an error because the second argument is never used 
    so it's never evaluated.

1.  See [infix](#infix-functions) and 
    [replacement functions](#replacement-functions).

1.  You use `on.exit()`; see [on exit](#on-exit) for details.