---
title: "Advanced R - Chapter 5 - Object Oriented Field Guide"
author: "Matthew Strimas-Mackey"
date: "December 5, 2014"
output:
  html_document:
    keep_md: yes
------

# OO field guide

##### Quiz

Think you know this material already? If you can answer the following questions correctly, you can safely skip this chapter. Find the answers at the end of the chapter in [answers](#oo-answers).

1. How do you tell what OO system (base, S3, S4, or RC) an object is 
   associated with?

1. How do you determine the base type (like integer or list) of an object?

1. What is a generic function?

1. What are the main differences between S3 and S4? What are the main 
   differences between S4 & RC?
   
## Base types

Underlying every R object is a C structure (or struct) that describes how that object is stored in memory. The struct includes the contents of the object, the information needed for memory management, and, most importantly for this section, a __type__. This is the __base type__ of an R object. 

[Data structures](#data-structures) explains the most common base types (atomic vectors and lists), but base types also encompass functions, environments, and other more exotic objects likes names, calls, and promises that you'll learn about later in the book. You can determine an object's base type with `typeof()`.

Functions that behave differently for different base types are almost always written in C, where dispatch occurs using switch statements (e.g., `switch(TYPEOF(x))`). Even if you never write C code, it's important to understand base types because everything else is built on top of them: S3 objects can be built on top of any base type, S4 objects use a special base type, and RC objects are a combination of S4 and environments (another base type). To see if an object is a pure base type, i.e., it doesn't also have S3, S4, or RC behaviour, check that `is.object(x)` returns `FALSE`.

## S3 

S3 is R's first and simplest OO system. It is an adhoc, informal system implemented primarily using functions. Objects, usually lists or atomic vectors, are assigned to a __class__ by setting the class attribute. Class is a character vector specifying the inheritance structure from least to most specific:

```{r}
x <- 'cat'
(class(x) <- c('cat', 'pet'))
y <- structure('dog', class=c('dog', 'pet'))
```

Unlike most OO systems, in which methods belong to objects, in S3 __methods__ belong to functions, called __generic functions__, which are responsible for __method dispatch__. Given an object, the generic function calls ```UseMethod()``` to dispatch to the correct S3 method. S3 methods have the standard naming convention: ```generic.class()```. 

```{r}
# Generic function calls UseMethod, with generic name
sound <- function(x) UseMethod("sound") # UseMethod dispatches on first argument of generic

# S3 methods invoked by generic
sound.cat <- function(x) 'meow'
sound.dog <- function(x) 'bark'
sound.pet <- function(x) 'feed me'

# The correct S3 method is called based on class
sound(x)
sound(y)
```

Inheritance works by progressing through the character vector of classes for until a matching S3 method is found. Evaluation may be explicitely passed to the second element of the class vector use ```NextMethod()```. Finally, a default method can be defined using ```generic.default()```:

```{r}
# No S3 method sound.fish exists, so sound.pet is called instead
z <- structure('fish', class=c('fish', 'pet'))
sound(z)

# Explicitely pass to parent class
sound.snake <- function(x) NextMethod()
s <- structure('snake', class=c('snake', 'pet'))
sound(s)

# Default method
sound.default <- function(x) 'who am I?'
sound('unknown')
```

Most S3 classes provide a constructor function, which has the same name as the class and ensures that the object is being created with the correct components:

```{r}
foo <- function(x) {
  if (!is.numeric(x)) stop("X must be numeric")
  structure(list(x), class = "foo")
}
foo('a')
foo(2)
```

### Recognising objects, generic functions, and methods

The functions `pryr::otype()` and `pryr::ftype()` return the type of the given object or function, respectively. ```methods()``` can be use to determine the set of S3 methods belonging to a given generic function or all the S3 methods for a given class. ```getS3method()``` returns the S3 method for a given generic and class:

```{r, message = FALSE}
library(pryr)

# Object type
df <- data.frame(x = 1:10, y = letters[1:10])
otype(df)    # A data frame is an S3 class
otype(df$x)  # A numeric vector isn't
otype(df$y)  # A factor is

# Function type
ftype(mean)
ftype(sum)
ftype(sound)
ftype(sound.dog)

# Methods belonging to given generic
methods(sound)

# S3 methods for a given class
methods(class='data.frame')

# S3 for given generic and class
getS3method('t', 'data.frame')
```

### Exercises

1.  Read the source code for `t()` and `t.test()` and confirm that 
    `t.test()` is an S3 generic and not an S3 method. What happens if 
    you create an object with class `test` and call `t()` with it?
    
    ```{r}
    body(t) # Call to UseMethod => generic
    body(t.test) # Call to UseMethod => generic

    ttest <- structure(list(), class='test')
    t(ttest) # fails
    methods(class='test') # no associated methods
    ```

1.  What classes have a method for the `Math` group generic in base R? Read 
    the source code. How do the methods work?
    
    ```{r}
    # What classes have methods for generic Math
    methods('Math')
    # > [1] Math.data.frame Math.Date       Math.difftime   Math.factor     Math.POSIXt
    body(Math.data.frame)
    ```
    
    For data frames, generics in the Math group are applied row-wise. The variable .Generic 
    refers to the actual generic being called

1.  R has two classes for representing date time data, `POSIXct` and 
    `POSIXlt`, which both inherit from `POSIXt`. Which generics have 
    different behaviours for the two classes? Which generics share the same
    behaviour?
    
    ```{r}
    str(lapply(methods(class='POSIXct'), getAnywhere)[[1]]$where)
    methods(class='POSIXlt')
    ```

1.  Which base generic has the greatest number of defined methods?

    ```{r}
    sapply(Filter(function(x) isGeneric(x) & !isS4(x), 
                  ls(all.names=TRUE, env = baseenv())), 
           function(x) length(methods(x)))
    ```
    
    * | has the most methods

1.  `UseMethod()` calls methods in a special way. Predict what the following
     code will return, then run it and read the help for `UseMethod()` to 
    figure out what's going on. Write down the rules in the simplest form
    possible.

    ```{r, eval = FALSE}
    y <- 1
    g <- function(x) {
      y <- 2
      UseMethod("g")
    }
    g.numeric <- function(x) y
    g(10) # Returns 2 NOT 1

    h <- function(x) {
      x <- 10
      UseMethod("h")
    }
    h.character <- function(x) paste("char", x)
    h.numeric <- function(x) paste("num", x)

    h("a") # Calls h.character, i.e. changing x to numeric ignored by UseMethod
    ```
    
    *   UseMethod dispatches on the first argument of the generic function. This object is passed directly
        to S3 method, ignoring any changes made to it inside the generic before the UseMethod call.
    *   The S3 method that UseMethod dispatches to is run within the evaluation environment of the 
        generic.

1.  Internal generics don't dispatch on the implicit class of base types.
    Carefully read `?"internal generic"` to determine why the length of `f` 
    and `g` is different in the example below. What function helps 
    distinguish between the behaviour of `f` and `g`?

    ```{r, eval = FALSE}
    f <- function() 1
    g <- function() 2
    class(g) <- "function"
    
    class(f)
    class(g)

    length.function <- function(x) "function"
    length(f)
    length(g)
    ```
    
    *   `f` is a primitive with `mode(f) = 'function'`, while `g` is an object of class 'function'.
        the `length` internal generic dispatches on the mode for `f` and the class for `g`.

## S4 {#s4}

S4 works in a similar way to S3, but it adds formality and rigour. Methods still belong to functions, not classes, but: 

* Classes have formal definitions which describe their fields and
  inheritance structures (parent classes).

* Method dispatch can be based on multiple arguments to a generic function,
  not just one.

* There is a special operator, `@`, for extracting slots (aka fields)
  from an S4 object.

### Recognising objects, generic functions, and methods

Recognising S4 objects, generics, and methods is easy. You can identify an S4 object because `str()` describes it as a "formal" class, `isS4()` returns `TRUE`, and `pryr::otype()` returns "S4". S4 generics and methods are also easy to identify because they are S4 objects with well defined classes.

```{r}
library(sp)

# Create an S4 object
data(meuse.grid)
m <- meuse.grid
coordinates(m) <- ~x+y

# An S4 object
isS4(m)
otype(m)

# An S4 generic
isS4(coordinates)
ftype(coordinates)

# Retrieve an S4 method, described later
sp_coordinates <- method_from_call(coordinates(m))
isS4(sp_coordinates)
ftype(sp_coordinates)
```

Use `is()` with one argument to list all classes that an object inherits from. Use `is()` with two arguments to test if an object inherits from a specific class.

```{r}
is(m)
is(m, "Spatial")
```

You can get a list of all S4 generics with `getGenerics()`, and a list of all S4 classes with `getClasses()`. This list includes shim classes for S3 classes and base types. You can list all S4 methods with `showMethods()`, optionally restricting selection either by `generic` or by `class` (or both). It's also a good idea to supply `where = search()` to restrict the search to methods available in the global environment.

### Defining classes and creating objects

In S3, you can turn any object into an object of a particular class just by setting the class attribute. S4 is much stricter: you must define the representation of a class with `setClass()`, and create an new object with `new()`. You can find the documentation for a class with a special syntax: `class?className`, e.g., `class?SpatialPoints`. 

An S4 class has three key properties:

* A __name__: an alpha-numeric class identifier. By convention, S4 class names
  use UpperCamelCase.

* A named list of __slots__ (fields), which defines slot names and
  permitted classes. For example, a person class might be represented by a
  character name and a numeric age: `list(name = "character", age = "numeric")`.
  \index{slots}

* A string giving the class it inherits from, or, in S4 terminology,
  that it __contains__. You can provide multiple classes for multiple
  inheritance, but this is an advanced technique which adds much
  complexity.

In `slots` and `contains` you can use S4 classes, S3 classes registered
with `setOldClass()`, or the implicit class of a base type. In `slots`
you can also use the special class `ANY` which does not restrict the input.

The following example creates a Person class with fields name and age, and an Employee class that inherits from Person. The Employee class inherits the slots and methods from the Person, and adds an additional slot, boss. To create objects we call `new()` with the name of the class, and name-value pairs of slot values. 

```{r}
setClass("Person",
  slots = list(name = "character", age = "numeric"))
setClass("Employee",
  slots = list(boss = "Person"),
  contains = "Person")

alice <- new("Person", name = "Alice", age = 40)
john <- new("Employee", name = "John", age = 20, boss = alice)
```

Most S4 classes also come with a constructor function with the same name as the class: if that exists, use it instead of calling `new()` directly.

To access slots of an S4 object use `@` or `slot()`: 

```{r}
alice@age
slot(john, "boss")
```

(`@` is equivalent to `$`, and `slot()` to `[[`.)

If an S4 object contains (inherits from) an S3 class or a base type, it will have a special `.Data` slot which contains the underlying base type or S3 object: \indexc{.Data}

```{r}
setClass("RangedNumeric",
  contains = "numeric",
  slots = list(min = "numeric", max = "numeric"))
rn <- new("RangedNumeric", 1:10, min = 1, max = 10)
rn@min
rn@.Data
```

### Creating new methods and generics

S4 provides special functions for creating new generics and methods. `setGeneric()` creates a new generic or converts an existing function into a generic. `setMethod()` takes the name of the generic, the classes the method should be associated with, and a function that implements the method. For example, we could take `union()`, which usually just works on vectors, and make it work with data frames: 

```{r}
# Create a new generic with the existing function union as the default
setGeneric("union")
# Create a formal method for the given generic and list of classes 
setMethod("union",
  signature=c(x = "data.frame", y = "data.frame"),
  function(x, y) {
    unique(rbind(x, y))
  }
)
```

If you create a new generic from scratch, you need to supply a function that calls `standardGeneric()`:

```{r}
setGeneric("myGeneric", function(x) {
  standardGeneric("myGeneric")
})
```

`standardGeneric()` is the S4 equivalent to `UseMethod()`.

### Method dispatch

If an S4 generic dispatches on a single class with a single parent, then S4 method dispatch is the same as S3 dispatch. The main difference is how you set up default values: S4 uses the special class `ANY` to match any class and "missing" to match a missing argument. Like S3, S4 also has group generics, documented in `?S4groupGeneric`, and a way to call the "parent" method, `callNextMethod()`. 

Method dispatch becomes considerably more complicated if you dispatch on multiple arguments, or if your classes use multiple inheritance. The rules are described in `?Methods`, but they are complicated and it's difficult to predict which method will be called. For this reason, I strongly recommend avoiding multiple inheritance and multiple dispatch unless absolutely necessary.

Finally, there are two methods that find which method gets called given the specification of a generic call:

```{r, eval = FALSE}
# From methods: takes generic name and class names
selectMethod("coordinates", list("SpatialPoints"))

# From pryr: takes an unevaluated function call
method_from_call(coordinates(m))
```

### Exercises

1.  Which S4 generic has the most methods defined for it? Which S4 class 
    has the most methods associated with it?

1.  What happens if you define a new S4 class that doesn't "contain" an 
    existing class?  (Hint: read about virtual classes in `?Classes`.)

1.  What happens if you pass an S4 object to an S3 generic? What happens 
    if you pass an S3 object to an S4 generic? (Hint: read `?setOldClass` 
    for the second case.)

## RC {#rc}

Reference classes (or RC for short) are the newest OO system in R. They were introduced in version 2.12. They are fundamentally different to S3 and S4 because: \index{RC} \index{reference classes|see{RC}} \index{objects!RC|see{RC}}

* RC methods belong to objects, not functions

* RC objects are mutable: the usual R copy-on-modify semantics do not apply

These properties make RC objects behave more like objects do in most other programming languages, e.g., Python, Ruby, Java, and C#. Reference classes are implemented using R code: they are a special S4 class that wraps around an environment.

### Defining classes and creating objects

Since there aren't any reference classes provided by the base R packages, we'll start by creating one. RC classes are best used for describing stateful objects, objects that change over time, so we'll create a simple class to model a bank account. \index{RC!classes} \index{classes!RC}

Creating a new RC class is similar to creating a new S4 class, but you use `setRefClass()` instead of `setClass()`. The first, and only required argument, is an alphanumeric __name__. While you can use `new()` to create new RC objects, it's good style to use the object returned by `setRefClass()` to generate new objects. (You can also do that with S4 classes, but it's less common.) \indexc{setRefClass()}

```{r}
Account <- setRefClass("Account")
Account$new()
```

`setRefClass()` also accepts a list of name-class pairs that define class __fields__ (equivalent to S4 slots). Additional named arguments passed to `new()` will set initial values of the fields. You can get and set field values with `$`: \index{fields}

```{r}
Account <- setRefClass("Account",
  fields = list(balance = "numeric"))

a <- Account$new(balance = 100)
a$balance
a$balance <- 200
a$balance
```

Instead of supplying a class name for the field, you can provide a single argument function which will act as an accessor method. This allows you to add custom behaviour when getting or setting a field. See `?setRefClass` for more details.

Note that RC objects are __mutable__, i.e., they have reference semantics, and are not copied-on-modify: \index{copy-on-modify!exceptions}

```{r}
b <- a
b$balance
a$balance <- 0
b$balance
```

For this reason, RC objects come with a `copy()` method that allow you to make a copy of the object:

```{r}
c <- a$copy()
c$balance
a$balance <- 100
c$balance
```

An object is not very useful without some behaviour defined by __methods__. RC methods are associated with a class and can modify its fields in place. In the following example, note that you access the value of fields with their name, and modify them with `<<-`. You'll learn more about `<<-` in [Environments](#binding). \index{RC!methods} \index{methods!RC} \indexc{<<-}

```{r}
Account <- setRefClass("Account",
  fields = list(balance = "numeric"),
  methods = list(
    withdraw = function(x) {
      balance <<- balance - x
    },
    deposit = function(x) {
      balance <<- balance + x
    }
  )
)
```

You call an RC method in the same way as you access a field:

```{r}
a <- Account$new(balance = 100)
a$deposit(100)
a$balance
```

The final important argument to `setRefClass()` is `contains`. This is the name of the parent RC class to inherit behaviour from. The following example creates a new type of bank account that returns an error preventing the balance from going below 0.

```{r, error = TRUE}
NoOverdraft <- setRefClass("NoOverdraft",
  contains = "Account",
  methods = list(
    withdraw = function(x) {
      if (balance < x) stop("Not enough money")
      balance <<- balance - x
    }
  )
)
accountJohn <- NoOverdraft$new(balance = 100)
accountJohn$deposit(50)
accountJohn$balance
accountJohn$withdraw(200)
```

All reference classes eventually inherit from `envRefClass`. It provides useful methods like `copy()` (shown above), `callSuper()` (to call the parent field), `field()` (to get the value of a field given its name), `export()` (equivalent to `as()`), and `show()` (overridden to control printing). See the inheritance section in `setRefClass()` for more details.

### Recognising objects and methods

You can recognise RC objects because they are S4 objects (`isS4(x)`) that inherit from "refClass" (`is(x, "refClass")`). `pryr::otype()` will return "RC".  RC methods are also S4 objects, with class `refMethodDef`.

### Method dispatch

Method dispatch is very simple in RC because methods are associated with classes, not functions. When you call `x$f()`, R will look for a method f in the class of x, then in its parent, then its parent's parent, and so on. From within a method, you can call the parent method directly with `callSuper(...)`. \index{RC!method dispatch rules}

### Exercises

1.  Use a field function to prevent the account balance from being directly
    manipulated. (Hint: create a "hidden" `.balance` field, and read the 
    help for the fields argument in `setRefClass()`.)

1.  I claimed that there aren't any RC classes in base R, but that was a 
    bit of a simplification. Use `getClasses()` and find which classes 
    `extend()` from `envRefClass`. What are the classes used for? (Hint: 
    recall how to look up the documentation for a class.)

## Picking a system {#picking-a-system}

Three OO systems is a lot for one language, but for most R programming, S3 suffices. In R you usually create fairly simple objects and methods for pre-existing generic functions like `print()`, `summary()`, and `plot()`. S3 is well suited to this task, and the majority of OO code that I have written in R is S3. S3 is a little quirky, but it gets the job done with a minimum of code. \index{objects!which system?}

```{r, eval = FALSE, echo = FALSE}
packageVersion("Matrix")

library(Matrix)
gs <- getGenerics("package:Matrix")
sum(gs@package == "Matrix")

length(getClasses("package:Matrix", FALSE))
```

If you are creating more complicated systems of interrelated objects, S4 may be more appropriate. A good example is the `Matrix` package by Douglas Bates and Martin Maechler. It is designed to efficiently store and compute with many different types of sparse matrices. As of version 1.1.3, it defines 102 classes and 20 generic functions. The package is well written and well commented, and the accompanying vignette (`vignette("Intro2Matrix", package = "Matrix")`) gives a good overview of the structure of the package. S4 is also used extensively by Bioconductor packages, which need to model complicated interrelationships between biological objects. Bioconductor provides many [good resources](https://www.google.com/search?q=bioconductor+s4) for learning S4. If you've mastered S3, S4 is relatively easy to pick up; the ideas are all the same, it is just more formal, more strict, and more verbose.

If you've programmed in a mainstream OO language, RC will seem very natural. But because they can introduce side effects through mutable state, they are harder to understand. For example, when you usually call `f(a, b)` in R you can assume that `a` and `b` will not be modified. But if `a` and `b` are RC objects, they might be modified in the place. Generally, when using RC objects you want to minimise side effects as much as possible, and use them only where mutable states are absolutely required. The majority of functions should still be "functional", and free of side effects. This makes code easier to reason about and easier for other R programmers to understand.

## Quiz answers {#oo-answers}

1.  To determine the OO system of an object, you use a process of elimination.
    If `!is.object(x)`, it's a base object. If `!isS4(x)`, it's S3. If 
    `!is(x, "refClass")`, it's S4; otherwise it's RC.
    
1.  Use `typeof()` to determine the base class of an object.

1.  A generic function calls specific methods depending on the class of 
    it inputs. In S3 and S4 object systems, methods belong to generic 
    functions, not classes like in other programming languages.
    
1.  S4 is more formal than S3, and supports multiple inheritance and
    multiple dispatch. RC objects have reference semantics, and methods 
    belong to classes, not functions.
    